Annotated example of index standardization with VAST

### 2022-05-05 doc/appendix-pcod-VAST-poisson-link.Rmd
Add draft poisson-link comparison

settings <- make_settings(
  n_x = 205, # number of vertices in the SPDE mesh
  Region = "User",
  purpose = "index2", # index of abundance with Gamma for positive catches
  fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
  zone = 9,
  FieldConfig = FieldConfig,
  RhoConfig = RhoConfig,
  ObsModel = c(2, 0), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
  bias.correct = FALSE,
  use_anisotropy = FALSE,
  max_cells = Inf, # use all grid cells from the extrapolation grid
  knot_method = "grid" # or "samples"
)
```

@@

```{r fit, echo = TRUE, results = 'hide'}
 # create folder for saved output:
dir.create(paste0(here("doc", "appendix-VAST")), showWarnings = FALSE)

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- 1
pcod <- as.data.frame(pcod) # ensure not a tibble
f <- here("doc", "appendix-VAST", "vast-cache.rds")
if (!file.exists(f)) {
  tictoc::tic()
  fit <- fit_model(
    settings = settings,
    Lat_i = pcod[, "lat"],
    Lon_i = pcod[, "lon"],
    t_i = pcod[, "year"],
    b_i = pcod[, "density"],
    a_i = pcod[, "effort"],
    input_grid = input_grid,
    working_dir = paste0(here("doc", "appendix-VAST"), "/")
  )
  tictoc::toc()
  saveRDS(fit, file = f)
} else {
  fit <- readRDS(f)
}
```

We can look at parameter estimates and their standard errors. First we see estimates from the binomial component and second we see estimates from the positive Gamma component.

@@

```{r sdmTMB-example, echo = TRUE, cache=TRUE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"),
  mesh = fit$spatial_list$MeshList$isotropic_mesh)

tictoc::tic()
fit_sdmTMB <- sdmTMB(
  density ~ 0 + as.factor(year), 
  data = pcod, mesh = mesh, 
  family = delta_gamma(),
  time = "year", silent = FALSE,
  control = sdmTMBcontrol(newton_loops = 1)
)
tictoc::toc()
fit_sdmTMB
```

@@

```{r save-plots, echo = TRUE, warning = FALSE, cache=TRUE}
plot(
  fit,
  check_residuals = FALSE,
  working_dir = paste0(here("doc", "appendix-VAST"), "/")
)
```

@@

```{r plot-knots, echo = TRUE, out.width="6in"}
knitr::include_graphics(here("doc", "appendix-VAST", "Data_and_knots.png"))
```

@@

```{r plot-density, echo = TRUE, out.width="6in"}
knitr::include_graphics(here("doc", "appendix-VAST", "ln_density-predicted.png"))
```

@@

```{r plot-index, echo = TRUE, out.width="3in"}
if (file.exists(here("doc", "appendix-VAST", "Index-Biomass.png"))) {
  knitr::include_graphics(here("doc", "appendix-VAST", "Index-Biomass.png"))
}
if (file.exists(here("doc", "appendix-VAST", "Index.png"))) {
  knitr::include_graphics(here("doc", "appendix-VAST", "Index.png"))
}
```

@@

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
tictoc::tic()
p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE,
  area = rep(4, nrow(qcs_grid))
)
ind <- get_index(p, bias_correct = FALSE)
tictoc::toc()
```

@@

```{r index-compare, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
vast_i <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
  mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
    se = Std..Error.for.ln.Estimate.) %>% 
  select(index, year, est, se) %>% 
  mutate(lwr = exp(log(est) + qnorm(0.025) * se)) %>% 
  mutate(upr = exp(log(est) + qnorm(0.975) * se))
sdm_i <- ind %>% mutate(index = "sdmTMB")
both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
  geom_ribbon(alpha = 0.1) +
  geom_line(alpha = 0.8) + 
  ylim(0, max(both_i$upr)) +
  coord_cartesian(expand = FALSE)
```

>>>>

settings <- make_settings(
  n_x = 205, # number of vertices in the SPDE mesh
  Region = "User",
  purpose = "index2", # index of abundance with Gamma for positive catches
  fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
  zone = 9,
  FieldConfig = FieldConfig,
  RhoConfig = RhoConfig,
  ObsModel = c(2, 1), # PG delta; c(10, 2) for Tweedie
  bias.correct = FALSE,
  use_anisotropy = FALSE,
  max_cells = Inf, # use all grid cells from the extrapolation grid
  knot_method = "grid" # or "samples"
)
```

@@

```{r fit, echo = TRUE, results = 'hide'}
 # create folder for saved output:
dir.create(paste0(here("doc", "appendix-VAST-poisson-link")), showWarnings = FALSE)

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- 1
pcod <- as.data.frame(pcod) # ensure not a tibble
f <- here("doc", "appendix-VAST-poisson-link", "vast-cache.rds")
if (!file.exists(f)) {
  tictoc::tic()
  fit <- fit_model(
    settings = settings,
    Lat_i = pcod[, "lat"],
    Lon_i = pcod[, "lon"],
    t_i = pcod[, "year"],
    b_i = pcod[, "density"],
    a_i = pcod[, "effort"],
    input_grid = input_grid,
    working_dir = paste0(here("doc", "appendix-VAST-poisson-link"), "/")
  )
  tictoc::toc()
  saveRDS(fit, file = f)
} else {
  fit <- readRDS(f)
}
```

We can look at parameter estimates and their standard errors.

@@

```{r sdmTMB-example, echo = TRUE, cache=TRUE}
mesh <- sdmTMB::make_mesh(pcod, xy_cols = c("X", "Y"),
  mesh = fit$spatial_list$MeshList$isotropic_mesh)

tictoc::tic()
fit_sdmTMB <- sdmTMB(
  density ~ 0 + as.factor(year), 
  data = pcod, mesh = mesh, 
  family = delta_poisson_link_gamma(),
  time = "year", silent = F,
  control = sdmTMBcontrol(newton_loops = 1)
)
tictoc::toc()
fit_sdmTMB
```

@@

```{r save-plots, echo = TRUE, warning = FALSE, cache=TRUE}
plot(
  fit,
  check_residuals = FALSE,
  working_dir = paste0(here("doc", "appendix-VAST-poisson-link"), "/")
)
```

@@

```{r plot-knots, echo = TRUE, out.width="6in"}
knitr::include_graphics(here("doc", "appendix-VAST-poisson-link", "Data_and_knots.png"))
```

@@

```{r plot-density, echo = TRUE, out.width="6in"}
knitr::include_graphics(here("doc", "appendix-VAST-poisson-link", "ln_density-predicted.png"))
```

@@

```{r plot-index, echo = TRUE, out.width="3in"}
if (file.exists(here("doc", "appendix-VAST-poisson-link", "Index-Biomass.png"))) {
  knitr::include_graphics(here("doc", "appendix-VAST-poisson-link", "Index-Biomass.png"))
}
if (file.exists(here("doc", "appendix-VAST-poisson-link", "Index.png"))) {
  knitr::include_graphics(here("doc", "appendix-VAST-poisson-link", "Index.png"))
}
```

@@

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
tictoc::tic()
p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE)

d <- p$data
n_i <- exp(d$est1)
w_i <- exp(d$est2)
p_i <- 1 - exp(-n_i)
r_i <- (n_i * w_i) / p_i
d_i <- p_i * r_i
d2_i <- exp(d$est1 + d$est2)
plot(d_i, d2_i)

library(dplyr)
nd <- qcs_grid
# nd <- pcod
nd$d_i <- d_i
ind_r <- group_by(nd, year) %>% 
  summarise(total = sum(d_i))
ind_r$total * 1

n_i <- exp(r$proj_eta[,1])
w_i <- exp(r$proj_eta[,2])
p_i <- 1 - exp(-n_i)
r_i <- (n_i * w_i) / p_i
d_i <- p_i * r_i
d2_i <- exp(d$est1 + d$est2)
nd$d_i <- d_i
ind_r_tmb <- group_by(nd, year) %>% 
  summarise(total = sum(d_i))
ind_r_tmb$total * 1

ind <- get_index(p, bias_correct = FALSE, area = 1)
tictoc::toc()
```

@@

```{r index-compare, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
vast_i <- read.csv(here("doc/appendix-VAST-poisson-link/Index.csv")) %>%
  mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
    se = Std..Error.for.ln.Estimate.) %>% 
  select(index, year, est, se) %>% 
  mutate(lwr = exp(log(est) + qnorm(0.025) * se)) %>% 
  mutate(upr = exp(log(est) + qnorm(0.975) * se))
sdm_i <- ind %>% mutate(index = "sdmTMB")
both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
  geom_ribbon(alpha = 0.1) +
  geom_line(alpha = 0.8) + 
  ylim(0, max(both_i$upr)) +
  coord_cartesian(expand = FALSE)
```
